Detecting diversifying selection for a trait from within and between-species genotypes and phenotypes

### 2024-07-03 manuscript/main-revision-2.tex
Revision round 2 manuscript files.

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\citep{simons_population_2018, sella_thinking_2019}.
Given this genetic architecture of traits, from two diverging populations, it is possible to distinguish which traits have evolved under natural selection in controlled experimental settings, by performing genetic cross between individuals~\citep{fraser_detecting_2020}.
Across several populations, by contrasting both trait differentiation (\Qst) and genetic differentiation (\Fst), so-called \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\citep{merila_comparison_2001, leinonen_comparative_2008}.
\Qst\ higher than \Fst\ is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations.
Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection~\citep{lamy_qst_2012}.
Other frameworks explicitly model genetic drift as a random process generating both trait and genetic differences between individuals and populations.
This integrated framework can discriminate between selection and genetic drift as a cause of trait differentiation between populations of the same species~\citep{ovaskainen_new_2011}.
However, regardless of the strengths and weaknesses of each method\citep{pujol_are_2008, edelaar_comparisons_2011, ovaskainen_new_2011}, tests of trait differentiation between populations are ultimately limited to recent local adaptation since they are based on the variation observed within a single species.

% Pair of species
To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
For example, starting from the same ancestral population, divergent lineages accumulate phenotypic changes that will reach fixation in the population.
These changes ultimately result in different mean trait values across lineages.
Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\citep{lande_genetic_1980, turelli_heritable_1984}.
Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\citep{khaitovich_neutral_2004}.
As an analogy, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \citet{mcdonald_adaptative_1991} test.
In such a test, inflation of divergence to the sister species is compared to polymorphism within species, while neutral makers (usually synonymous sites) are used to determine the neutral expectation and thus are used for normalization.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\citep{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Deviations from the assumptions of the Brownian process are however well known.
When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\citep{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also changing as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\citep{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Also, a better fit of a Brownian could be due to a trait evolving with a rate two low compared to the timespan on which it is measured~\citep{grabowski_cautionary_2023}.
Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\citep{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\citep{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\citep{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
Between to within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of nucleotide sequence variations at both scales.
It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
The ratio that we propose can be considered as a neutrality index for a any quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
From the field of population genetics, while \QstFst\ methods and their derivatives ultimately seek trait differentiation among different populations from the same species~\citep{pujol_are_2008, ovaskainen_new_2011}, our study can be seen as their macro-evolutionary generalization to account for phylogenetic relationships between species.
From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {figure1}
    \caption{
        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
        Within species, the genetic variance allows the estimation of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

@@

We denote $\EstNI$, $\EstRateBetween$ and $\EstRateWhithin$ the point estimates of respectively $\NI$, $\RateBetween$ and $\RateWhithin$.
For each species with data available, $\RateWhithin$ as defined in eq.~\ref{eq:def-rate-pheno-pop} can be seen as a replicate sample.
Thus, $\EstRateWhithin$ can be obtained by averaging out across all the sampled species.
On the other hand, $\RateBetween$ such as as defined in eq.~\ref{eq:def-rate-phy} only refers to a pair of species, and thus must be generalized to account for different species divergence, as is done in the comparative framework~\citep{felsenstein_phylogenies_1985, omeara_testing_2006}.
Generally, $\EstRateWhithin$ can thus be seen as an estimate of the rate of evolution of the quantitative trait along a phylogenetic tree, when the tree is measured in units of $4d$ ($d$ being the nucleotide divergence).
As such, any phylogenetic comparative methods that allow the estimation of phenotypic rates of evolution on a tree scaled by $4d$, instead of time as is usually the case, can be used to estimate $\EstRateWhithin$.
We provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.

@@

Finally, between-species variation $\EstRateBetween$ is estimated as~\citep{omeara_testing_2006}:
\begin{gather}
    \EstRateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}. \label{eq:estimated-rate-phy}
\end{gather}

For a given species $\Spi$ with inter-individual data available, additive genetic variance of a trait ($\VarGeneticSpi$) is the product of heritability ($\Heritability_{i}$) and phenotypic variance ($V_{\Trait, i}$).
The ratio of $\VarGeneticSpi$ over nucleotide diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RateWhithin$.
Averaged across all species, we obtain the estimate $\EstRateWhithin$ as:
\begin{gather}
    \EstRateWhithin = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heritability_{i}}{ \pi_{i}}. \label{eq:estimated-rate-pop}
\end{gather}

As depicted in fig.~\ref{fig:methods}, the neutrality index is estimated as:
\begin{gather}
    \EstNI = \frac{\EstRateBetween}{\EstRateWhithin}. \label{eq:estimated-NI}
\end{gather}

@@

In the previous section, $\EstNI$ is estimated independently for each trait of interest.
Here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree.
Trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\citep{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The branch lengths of the tree used to model the Brownian process runs is measured in units of $4d$ ($d$ being the nucleotide divergence).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait when measured in $4d$ units, and thus equate to $\EstRateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).

\subsubsection*{Bayesian estimate}
The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\EstNI$) for traits of interest.
We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\citep{latrille_inferring_2021}.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
We used Metropolis-Hastings algorithm to sample $\Brownian$, while the posterior distribution of $\CovarianceMatrix$ is sampled using Gibbs sampling.
For each trait and each species, the prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
Heritability and phenotypic variance for each trait are combined with nucleotide diversity to compute $\EstRateWhithin$ for each species before being averaged across species (as in eq.~\ref{eq:estimated-rate-pop}).
From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
The posterior distribution of $\EstNI$ thus allows testing for deviation from neutrality (Fig.~\ref{fig:methods}), for example, by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.

@@

Our method assumes that the narrow-sense heritability ($\Heritability$) of a trait is known such as to estimate additive genetic variance ($\VarGenetic$) from phenotypic variance ($\VarPhenotype$) as $\VarGenetic = \Heritability \Multiply \VarPhenotype$.
Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is underpowered.
Indeed, if the additive genetic variance is substituted by phenotypic variance, it is equivalent to assuming complete heritability ($\Heritability = 1$).
Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\EstNI$.
It is, however, possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
Additionally, empirical estimates of $\Heritability$ are surprisingly stable across species and fall within the range of 0.2-0.5 in a vast majority of phenotypic traits tested~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
Alternatively, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\EstNI$ since $\Heritability \leq H^2$.
If available, such prior knowledge on $\Heritability$ can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection.

In contrast to the test of diversifying selection, the test for stabilizing selection is invalid if $\EstNI$ is underestimated.
Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.

@@

At each generation, parents were randomly sampled with a weight proportional to their fitness ($W$).
Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\citep{tenaillon_utility_2014,blanquart_epistasis_2016}.
More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
Mutations were considered as a displacement of the phenotype in the multidimensional space.
Beneficial mutations moved the phenotype closer to the optimum, while deleterious mutations moved it further away.
Stabilizing selection was implemented by fixing the optimum phenotype to a single value ($\lambda=0.0$).
Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
Neutral evolution was implemented by fixing the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.

@@

We also analyzed a dataset of primate species, with the nucleotide variation obtained from \citet{kuderna_global_2023} and the quantitative trait variation also from \citet{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
As such, the evidence for $\EstNI > 1$ does not necessarily imply that the trait is evolving under diversifying selection since non-neutral markers included in the estimate of divergence can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).

@@

For a neutral trait, the genetic architecture, meaning the number of loci encoding the trait and the average effect of a mutation on the trait, is formally related to both within and between-species variation of the trait.
We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see Materials and Methods), suggesting that traits for which this relationship was not verified were putatively under selection.
Under stabilizing selection, the variation between species is depleted because the mean trait is maintained toward similar values between different species, which leads to $\NI < 1$.
In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\citep{hansen_stabilizing_1997}, which leads to $\NI > 1$.
Our neutrality index for a quantitative trait leveraged the data for any number of species, and took advantage of the signal over the whole phylogenetic tree, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (Fig.~\ref{fig:methods}).
This statistic was obtained as a maximum likelihood estimate ($\EstNI$), from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
We also devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilizing selection as $\proba [\EstNI < 1]$.

@@

The phenotype is encoded by a genetic architecture and is thus ultimately bounded.
At the macro-evolutionary scale, phenotypic divergence should plateau at some point, ultimately resulting in a $\RateBetween$.

>>>>

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\citep{simons_population_2018, sella_thinking_2019}.
Given this genetic architecture of traits, from two diverging populations, it is possible to distinguish which traits have evolved under natural selection in controlled experimental settings, by performing genetic cross between individuals~\citep{fraser_detecting_2020}.
Across several populations, by contrasting both trait differentiation (\Qst) and genetic differentiation (\Fst), so-called \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\citep{merila_comparison_2001, leinonen_comparative_2008}.
\Qst\ higher than \Fst\ is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations.
Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection~\citep{lamy_qst_2012}.
Other frameworks explicitly model genetic drift as a random process generating both trait and genetic differences between individuals and populations.
This integrated framework can discriminate between selection and genetic drift as a cause of trait differentiation between populations of the same species~\citep{ovaskainen_new_2011}.
However, regardless of the strengths and weaknesses of each method~\citep{pujol_are_2008, edelaar_comparisons_2011, ovaskainen_new_2011}, tests of trait differentiation between populations are ultimately limited to recent local adaptation since they are based on the variation observed within a single species.

% Pair of species
To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
For example, starting from the same ancestral population, divergent lineages accumulate phenotypic changes that will reach fixation in the population.
These changes ultimately result in different mean trait values across lineages.
Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\citep{lande_genetic_1980, turelli_heritable_1984, felsenstein_phylogenies_1988a}.
Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\citep{khaitovich_neutral_2004}.
As an analogy, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \citet{mcdonald_adaptative_1991} test.
In such a test, inflation of divergence to the sister species is compared to polymorphism within species, while neutral makers (usually synonymous sites) are used to determine the neutral expectation and thus are used for normalization.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\citep{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Deviations from the assumptions of the Brownian process are however well known.
When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\citep{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also changing as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\citep{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Second, a better fit of a Brownian could be due to a trait evolving with a rate too low compared to the timespan on which it is measured~\citep{grabowski_cautionary_2023}.
And third, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\citep{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\citep{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\citep{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
Across many species, within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within species variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of nucleotide sequence variations at both scales.
It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
The ratio that we propose can be considered as a neutrality index for any quantitative trait ~\citep{lynch_rate_1990}, while articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
From the field of population genetics, while \QstFst\ methods and their derivatives ultimately seek trait differentiation among different populations from the same species~\citep{pujol_are_2008, ovaskainen_new_2011}, our study can be seen as their macro-evolutionary analog to account for phylogenetic relationships between species.
From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {figure1}
    \caption{
        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\RateBetween$, which is normalized by nucleotide divergence.
        Within species, the genetic variance allows the estimation of within-species trait variation, $\RateWhithin$, which is  normalized by nucleotide diversity.
        $\NI$ is the ratio of $\RateBetween$ over $\RateWhithin$.
        Under neutral evolution, $\NI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\NI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\NI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

@@

We hereby seek to obtain point estimates of $\RateBetween$, $\RateWhithin$ and ultimately $\NI$.
For each species with data available, $\RateWhithin$ as defined in eq.~\ref{eq:def-rate-pheno-pop} can be seen as a replicate sample.
Thus, $\RateWhithin$ can be obtained by averaging out across all the sampled species.
On the other hand, $\RateBetween$ such as as defined in eq.~\ref{eq:def-rate-phy} only refers to a pair of species, and thus must be generalized to account for different species divergence, as is done in the comparative framework~\citep{felsenstein_phylogenies_1985, omeara_testing_2006}.
Generally, $\RateBetween$ can thus be seen as an estimate of the rate of evolution of the quantitative trait along a phylogenetic tree, when the tree is measured in units of $4d$ ($d$ being the nucleotide divergence).
As such, any phylogenetic comparative methods that allow the estimation of phenotypic rates of evolution on a tree scaled by $4d$, instead of time as is usually the case, can be used to estimate $\RateWhithin$.
We provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.

@@

Finally, between-species variation $\RateBetween$ is estimated as~\citep{omeara_testing_2006}:
\begin{gather}
    \RateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}. \label{eq:estimated-rate-phy}
\end{gather}

For a given species $\Spi$ with inter-individual data available, additive genetic variance of a trait ($\VarGeneticSpi$) is the product of heritability ($\Heritability_{i}$) and phenotypic variance ($V_{\Trait, i}$).
The ratio of $\VarGeneticSpi$ over nucleotide diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RateWhithin$.
Averaged across all species, we obtain the estimate $\RateWhithin$ as:
\begin{gather}
    \RateWhithin = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heritability_{i}}{ \pi_{i}}. \label{eq:estimated-rate-pop}
\end{gather}

As depicted in fig.~\ref{fig:methods}, the neutrality index is estimated as:
\begin{gather}
    \NI = \frac{\RateBetween}{\RateWhithin}. \label{eq:estimated-NI}
\end{gather}

@@

In the previous section, $\NI$ is estimated independently for each trait of interest.
Here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree, since simultaneously estimating all $\RateBetween$ allows improving their estimation~\citep{adams_multivariate_2018}.
More specifically, trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\citep{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The branch lengths of the tree used to model the Brownian process runs is measured in units of $4d$ ($d$ being the nucleotide divergence).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait when measured in $4d$ units, and thus equate to $\RateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
Of note, modeling trait evolution as a multi-dimensional process is reliable only if $\Ntrait \ll \NbrTaxa$, meaning that the number of species is largely superior to the number of traits~\citep{adams_multivariate_2018}.
Thus, relying on a $\Ntrait$-dimensional process should be reserved for a handful of allometric traits (e.g. brain mass and body mass).
If $\Ntrait$ is large, the traits are better tested independently each with a $1$-dimensional Brownian process, which is a specific case of the multi-dimensional process.

\subsubsection*{Bayesian estimate}
The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\NI$) for traits of interest.
We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\citep{latrille_inferring_2021}.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
We used Metropolis-Hastings algorithm to sample $\Brownian$, while the posterior distribution of $\CovarianceMatrix$ is sampled using Gibbs sampling.
For each trait and each species, the prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
Heritability and phenotypic variance for each trait are combined with nucleotide diversity to compute $\RateWhithin$ for each species before being averaged across species (as in eq.~\ref{eq:estimated-rate-pop}).
From $\RateWhithin$ estimated independently for each trait and the diagonal elements of $\CovarianceMatrix$ (i.e. the $\RateBetween$ for each trait), the posterior distribution of $\NI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
The posterior distribution of $\NI$ thus allows testing for deviation from neutrality (Fig.~\ref{fig:methods}), for example, by computing $\proba [\NI > 1 ]$ to test for evidence of diversifying selection and $\proba [\NI < 1 ]$ to test for evidence of stabilizing selection.

@@

Our method assumes that the narrow-sense heritability ($\Heritability$) of a trait is known such as to estimate additive genetic variance ($\VarGenetic$) from phenotypic variance ($\VarPhenotype$) as $\VarGenetic = \Heritability \Multiply \VarPhenotype$.
Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is underpowered.
Indeed, if the additive genetic variance is substituted by phenotypic variance, it is equivalent to assuming complete heritability ($\Heritability = 1$).
Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\NI$.
It is, however, possible to test for diversifying selection because testing for $\NI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\NI$ since $\Heritability \leq H^2$ and thus can be used to detect diversifying selection if $\Heritability$ is not available.
Additionally, empirical estimates of $\Heritability$ are surprisingly stable across species and fall within the range of 0.2-0.5 in a vast majority of phenotypic traits tested~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
Thus, if available, such prior knowledge on $\Heritability$ can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection.

In contrast to the test of diversifying selection, the test for stabilizing selection is invalid if $\NI$ is underestimated.
Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.

@@

At each generation, parents were randomly sampled with a weight proportional to their fitness ($W$).
Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\citep{tenaillon_utility_2014,blanquart_epistasis_2016}.
More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
Mutations were considered as a displacement of the phenotype in the multidimensional space.
Beneficial mutations moved the phenotype closer to the optimum, while deleterious mutations moved it further away.
Stabilizing selection was implemented by fixing the optimum phenotype to a single value ($\lambda=0.0$).
Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
Neutral evolution was implemented by flattening the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.

@@

We also analyzed a dataset of primate species, with the nucleotide variation obtained from \citet{kuderna_global_2023} and the quantitative trait variation also from \citet{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
As such, the evidence for $\NI > 1$ does not necessarily imply that the trait is evolving under diversifying selection since non-neutral markers included in the estimate of divergence can lead to a spurious $\NI > 1$ (see table~\ref{table:assumptions}).

@@

For a neutral trait, the genetic architecture, meaning the number of loci encoding the trait and the average effect of a mutation on the trait, is formally related to both within and between-species variation of the trait.
We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see Materials and Methods), suggesting that traits for which this relationship was not verified were putatively under selection.
Under stabilizing selection, the variation between species is depleted because the mean trait is maintained toward similar values between different species, which leads to $\NI < 1$.
In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\citep{hansen_stabilizing_1997}, which leads to $\NI > 1$.
Our neutrality index for a quantitative trait leveraged the data for any number of species, and took advantage of the signal over the whole phylogenetic tree, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (Fig.~\ref{fig:methods}).
This statistic was obtained as a maximum likelihood estimate from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
We also devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\NI > 1]$, and stabilizing selection as $\proba [\NI < 1]$.

@@

In our simulation setting, the genetic architecture underlying the phenotype is not changing along the phylogeny.
As a result, the phenotype may be bounded by what such genetic architecture can produce, and this could cause a slowdown of phenotypic divergence over time, ultimately resulting in a decreased $\RateBetween$.
